---
title: "R PROJECT"
subtitle: "A datadriven analysis of brain tumors"
author: 
  - "Marco Käferbeck"
  - "Sam Corvin"
  - "Tobias Granegger"
  - "Fabian Veis"
format: 
  revealjs:
    theme: default
    slide-number: true
    toc: true
    toc-depth: 2
    logo: LogoWU.png
    transition: slide
    chalkboard: true
    #highlight-style: github
    fig-align: center
    code-line-numbers: true
    code-fold: show
    footer: "Data Literacy – WU Vienna"
    progress: true
    layout: true
    #title-slide-attributes:
      #data-background-color: "#002b36"
      #data-background-size: cover
editor: visual
---

## Data set (I/II)

**Brain Tumor Dataset:** 

-   Dataset contains simulated data for brain tumor diagnosis, treatment, and patient details
-   20 columns and 20,000 rows
-   Provides information such as patient demographics, tumor characteristics, symptoms, treatment details, and follow-up requirements

## Data set (II/II)

Summary information:

```{r}
library(readr)
data <- read_csv("brain_tumor_dataset.csv")

n_patients <- nrow(data)
avg_age <- round(mean(data$Age), 1)
gender_dist <- prop.table(table(data$Gender))
tumor_type_dist <- prop.table(table(data$Tumor_Type))
common_locations <- names(sort(table(data$Location), decreasing = TRUE)[1:3])
top_histologies <- names(sort(table(data$Histology), decreasing = TRUE)[1:3])
radiation_pct <- round(mean(data$Radiation_Treatment == "Yes") * 100)
chemo_pct <- round(mean(data$Chemotherapy == "Yes") * 100)
surgery_pct <- round(mean(data$Surgery_Performed == "Yes") * 100)
avg_survival <- round(mean(data$Survival_Rate), 1)
avg_growth <- round(mean(data$Tumor_Growth_Rate), 2)

cat(glue::glue("
- Patients: {n_patients}
- Average Age: {avg_age} years
- Gender: {round(gender_dist['Male']*100)}% Male, {round(gender_dist['Female']*100)}% Female
- Tumor Types: {round(tumor_type_dist['Malignant']*100)}% Malignant, {round(tumor_type_dist['Benign']*100)}% Benign
- Common Locations: {paste(common_locations, collapse = ', ')}
- Top Histology Types: {paste(top_histologies, collapse = ', ')}
- Treatment:
   - Radiation: {radiation_pct}%
   - Chemotherapy: {chemo_pct}%
   - Surgery: {surgery_pct}%
- Mean Survival Rate: {avg_survival}%
- Mean Tumor Growth Rate: {avg_growth}
"))
```

## Research question

What are the most significant clinical and demographic factors influencing the survival rate of brain tumor patients?

How does the most influential factor (e.g. tumor stage or treatment response) causally affect the survival rate of brain tumor patients?

## Exploratory data analysis

```{r}
# Datentypen vorbereiten
data <- data %>%
  mutate(across(where(is.character), as.factor))

data$Age <- as.numeric(data$Age)
hist(data$Age, main = "Age distribution", xlab = "Alter", col = "skyblue", breaks = 10)
```

## Random forest, feature of importance

## Causal method, machine learning, direct acyclical analysis

## Conclusion
